### 2015-03-21 src/paper3/p3_methods.tex
Updated papers
Processed Luc and Pieters comments on Intro, P3 and P4

\section{Materials and methods}

@@

A motorized linear sled (see \cite{clemens2012} for details) was used to laterally translate participants following a minimum jerk profile of fixed duration (1 \si{\second}) and amplitudes ranging from 1 to 27 \si{\centi\metre}. Participants were seated on the sled such that the inter-aural axis aligned with the motion axis. They were restrained using a five-point seat belt and a chin rest. In addition, the head was held in place using an ear-fixed mold. Auditory cues were suppressed using white noise presented through in-ear headphones. Experiments were conducted in complete darkness except for visual fixation points, projected by a laser pointer on a black bar 50cm in front of the participant at eye level. Laser pointers used to project body-fixed targets were attached to the sled. Those used to project world-fixed targets were mounted on the wall behind the sled.

Eye movements were recorded at 500 \si{\hertz} using an EyeLink II (SR Research, Kanata, Canada) system whose cameras were mounted to the sled and therefore remained stable with respect to the head during the entire experiment. Because the head and body positions were fixed during the experiment, the orientation of the eyes within the head, as measured by the tracker, was equivalent to the orientation of the eyes in space. The eye tracking system was calibrated before each session using 11 evenly spaced calibration points ranging from -22 to \siang{22}. We used linear regression to link EyeLink measurements to gaze angles.

@@

\begin{figure}
    \includegraphics[width=1.0\textwidth]{src/paper3/figure1.eps}

@@

The time evolution of a single trial is shown in \figref{p3:fig1}. Each trial started with the onset of a central fixation point (i.e. aligned between the eyes) for 0.5 \si{\second}. Subsequently, the first translation interval commenced.  Depending on the fixation type, the fixation point remained visible (world and body) or was extinguished (free) during the translation interval. The trial shown in the figure depicts the reference translation of 10 \si{\centi\metre} with world fixation. After this first interval, a delay followed in which the participant was kept in complete darkness for 1.75 \si{\second}. Then, the central fixation point reappeared, followed 0.5 \si{\second} later by the second interval, in which the probe translation was presented with amplitudes in the range of 1 to 27 \si{\centi\metre}. The fixation type in the probe interval was always different than in the associated reference interval (the trial in \figref{p3:fig1} illustrates body fixation). After the second interval, the participant had to indicate using a joystick whether he or she perceived the second translation as longer or shorter than the first.

@@

Thus, a trial consists of two translations with different fixation types; in the three main conditions we compare the body versus world, world versus free, and body versus free fixation types. For each main condition, we varied which fixation type served as the reference stimulus and the order in which reference and probe were presented, which gives a total of four variations per main condition (see \tabref{p3:tab1}). In addition we varied translation direction (either leftward or rightward on consecutive trials). The amplitude of the probe translation was adaptively chosen based on the participants' earlier responses \cite{kontsevich1999} to determine the point of subjective equality. This was done separately for all 24 trial types (3 main conditions x 2 reference stimuli x 2 reference/probe orders x 2 translation directions; see \tabref{p3:tab1}). A total of 25 trials were collected per trial type yielding a total of 200 trials for each of the three main conditions.

Trials were presented in three one-hour sessions. To prevent dark adaptation, we turned on the lights for 5 \si{\second} after every block of 6 trials, and for at least 30 seconds every 4 blocks. We made sure that each of the 24 unique trial types were presented once every 4 blocks. After each block, the adaptive procedure determined which translation amplitudes to test in the following block. To increase the number of data-points available to the adaptive psychometric procedure at the beginning of the experiment, we collapsed across translation direction and reference order for the first 10 trials. After those blocks the procedure ran separately for each of the 24 distinct trial types.

@@

For each combination of the three main conditions, and the two reference/probe orders (see \tabref{p3:tab1}), we quantified the perceived probe translation by calculating the probability of the probe translation judged longer compared to the reference translation as a function of actual probe translation. We used a maximum likelihood fit of a cumulative Gaussian function to summarize the psychometric data:

@@

in which $|x|$ represents the size of the probe displacement. The mean of the Gaussian represents the point of subjective equality (PSE). The slope of the curve reflects the precision ($1/\sigma$) of reference-probe discrimination performance. Parameter $\lambda$, representing the lapse rate, accounts for stimulus-independent errors caused by subject lapses or mistakes and was restricted to small values ($\lambda < 0.06$). Fits were performed using the Psignifit toolbox \cite{wichmann2001,wichmann2001b}.

In order to elucidate the effects of eye movements on self-motion perception, we quantified eye movement during each sled movement. To this end, we first corrected eye movements for drift based on the initial fixations at the start of each trial when the sled was still stationary. Subsequently, these actual eye movements were normalized as a ratio ($g_i$) of the theoretical eye movements that would have occurred had the subject kept fixating the initial, world stationary, target location. Trials in which the normalized final eye position was more than two standard deviations away from the average normalized final eye position were excluded from further analysis. This was done separately for each subject and fixation type. Trials containing blinks were also removed. In total 12.6\% of all trials were rejected.

Using a simple cue integration model, we investigated whether inter-subject and inter-condition differences in the observed PSEs in conditions containing a translation under free fixation depend on actual eye movement behavior. For these analyses we used the normalized eye movements. Ideally, the normalized eye movement would be $g_i \approx 0$ for body fixation and $g_i \approx 1$ for world fixation. With free fixation, eye movements are expected to be in-between body and world, $0 < g_i < 1$, but the exact value is participant dependent. Variable $i$ represents either the reference ($r$) or probe ($p$) interval. We model perceived distance, $p_i$, as a weighted linear combination of a vestibular and an oculomotor estimate of translation (Eq. 2). We assume that the vestibular estimate is equal to the actual translation, $m_i$, and that the oculomotor estimate is equal to the actual translation scaled by the normalized eye movement, $g_i m_i$. We further assume that the weights sum to one, such that the weighting parameter $\alpha$ regulates the eye movement contribution and $1 - \alpha$ the vestibular contribution:

\begin{equation}
\label{p3:eq2}
p_i = g_i m_i \alpha + m_i (1 - \alpha)
\end{equation}

By definition, at the PSE the estimated translation of the probe displacement ($p_p$) is equal to the estimated translation of the reference displacement ($p_r$). By substituting both sides by the right hand side of \eqnref{p3:eq2} and using subscripts for reference ($r$) and probe translations ($p$), we obtain:

\begin{equation}
\label{p3:eq3}
g_r m_r \alpha + m_r (1 - \alpha) = g_p m_p \alpha + m_p (1 - \alpha)
\end{equation}

In this equation $m_r$ equals the actual reference translation of 10 cm and $m_p$ equals the measured PSE. After rewriting \eqnref{p3:eq3} into \eqnref{p3:eq4}, we performed linear regression to find the weight ($\alpha$) that minimizes the sum of squared errors ($\sum{\epsilon^2}$) over the body versus world and world versus body conditions only:

\begin{equation}
\label{p3:eq4}
m_p - m_r = \alpha ( g_r m_r - g_p m_r + m_p - m_r ) + \epsilon
\end{equation}

Next we used this subject specific $\alpha$ to predict the PSEs for the conditions containing a free-fixation. To this end we solved \eqnref{p3:eq3} for $m_p$ (\eqnref{p3:eq5}), which is an estimate ($\hat{m}_p$) of the PSE, and plugged in numerical values for the reference translation and the average normalized eye movements over the reference and probe trials.
                                                                                            \begin{equation}
\label{p3:eq5}
\hat{m}_p = \frac
	{g_r m_r \alpha + m_r (1 - \alpha)}
	{g_p \alpha + (1 - \alpha)}
\end{equation}

>>>>

\section{Materials and methods}
\label{p3:sec:methods}

@@

A motorized linear sled \cite<see>[for details]{clemens2012} was used to laterally translate participants following a minimum jerk profile \cite{flash1985} of fixed duration (1 \si{\second}) and amplitudes ranging from 1 to 27 \si{\centi\metre}. Participants were seated on the sled such that the inter-aural axis aligned with the motion axis. They were restrained using a five-point seat belt and a chin rest. In addition, the head was held in place using a sled-fixed mold which resembled head-phones and pressed down on the head surrounding the pinnea . Auditory cues were suppressed using white noise presented through in-ear headphones. Experiments were conducted in complete darkness except for visual fixation points, projected by a laser pointer on a black bar 50 \si{\centi\metre} in front of the participant at eye level. Laser pointers used to project body-fixed targets were attached to the sled. Those used to project world-fixed targets were mounted on the wall behind the sled.

Eye movements were recorded at 500\si{\hertz} using an EyeLink II (SR Research, Kanata, Canada) system whose cameras were mounted to the sled and therefore remained stable with respect to the head during the entire experiment. Because the head and body positions were fixed during the experiment, the orientation of the eyes within the head, as measured by the tracker, was equivalent to the orientation of the eyes in space. The eye tracking system was calibrated before each session using 11 evenly spaced calibration points ranging from -22 to \siang{22} degrees. We used linear regression to link EyeLink measurements to gaze angles.

@@

\begin{figure}
    \includegraphics[width=1.0\textwidth]{src/paper3/figure1.pdf}

@@

The time evolution of a single trial is shown in \figref{p3:fig1}. Each trial started with the onset of a central fixation point (i.e. aligned between the eyes) for 0.5 \si{\second}. Subsequently, the first translation interval commenced.  Depending on the fixation type, the fixation point remained visible (world and body) or was extinguished (free) during the translation interval. The trial shown in the figure depicts the 10 \si{\centi\metre} reference translation with world fixation. After this first interval, a delay followed in which the participant was kept in complete darkness for 1.75 \si{\second}. Then, the central fixation point reappeared, followed 0.5 \si{\second} later by the second interval, in which the probe translation was presented. The set of possible probe translations ranged from 1 to 27 \si{\centi\metre} in equidistant steps of 0.4 \si{\milli\metre}. The fixation type in the probe interval was always different than in the associated reference interval (the trial in \figref{p3:fig1} illustrates body fixation). After the second interval, the participant had to indicate whether he or she perceived the second translation as longer or shorter than the first using a 1-dimensional joystick. Moving the poke away from the body indicted that the second movement was longer, while moving it towards the body indicated that the second movement was shorter.

@@

Thus, a trial consists of two translations with different fixation types; in the three main conditions we compare the body versus world, world versus free, and body versus free fixation types. For each main condition, we varied which fixation type served as the reference stimulus and the order in which reference and probe were presented, which gives a total of four variations per main condition (see \tabref{p3:tab1}). In addition we varied translation direction (either leftward or rightward on consecutive trials). The amplitude of the probe translation was adaptively chosen using the Psi method. This method picks the amplitude for the next trial which maximizes the expected decrease in entropy based on participants’ responses to earlier trials \cite{kontsevich1999}. This was done separately for all 24 trial types (3 main conditions x 2 reference stimuli x 2 reference/probe orders x 2 translation directions; see \tabref{p3:tab1}). A total of 25 trials were collected per trial type yielding a total of 200 trials for each of the three main conditions.

Trials were presented in three one-hour sessions. To prevent dark adaptation, we turned on the lights for 5 \si{\second} after every block of 6 trials, and for at least 30 seconds every 4 blocks. We made sure that each of the 24 unique trial types were presented once every 4 blocks. After each block, the adaptive procedure determined which translation amplitudes to test in the following block. To increase the number of data-points available to the adaptive psychometric procedure at the beginning of the experiment, we collapsed across translation direction and reference order for the first 10 trials of every condition. After those collapsed trials, the procedure ran separately for each of the 24 distinct trial types.

@@

For each combination of the three main conditions, and  the two reference/probe orders (see \tabref{p3:tab1}), we quantified the perceived probe translation  by calculating the probability of the probe translation judged longer compared to the 10 \si{\centi\metre} reference translation as a function of actual probe translation, given by x. We used a maximum likelihood fit of a cumulative Gaussian function to summarize the psychometric data:

@@

in which $|x|$ represents the size of the absolute probe displacement. The mean of the Gaussian represents the point of subjective equality (PSE). The slope of the curve reflects the precision ($1/\sigma$) of reference-probe discrimination performance. Parameter $\lambda$, representing the lapse rate, accounts for stimulus-independent errors caused by subject lapses or mistakes and was restricted to small values ($\lambda < 0.06$. Fits were performed using the Psignifit toolbox \cite{wichmann2001,wichmann2001b}.

For each trial type (see \tabref{p3:tab1}), we also qualified eye movements, corrected for drift, based on initial fixation. The main source of drift were tiny lateral movements of the eye tracking cameras due to sled motion. We discarded trials containing blinks as well as trials in which the final eye position exceeded two standard deviations from the condition's average. Based on these criteria, 6.1\%, 3.6\% and 1.6\% of all trials were rejected based on errors in body, world, and free fixation respectively. In addition we rejected 1.2\% of all trials because participants blinked within the movement interval.

For the remaining trials, we computed the average ratio between the measured eye excursion, $\varphi_i$, and the angle that woud be needed were the trial testing the world-fixed condition. The latter is computed by taking the arc-tangent of the actual translation distance, $m_i$, divided by the fixation depth, $d_i$, which for small $\varphi$ can be approximated by $g = \varphi m/d$. We computed this ratio, $g$, for every fixation type and interval (see \tabref{p3:tab1}). Ideally, for body-fixed trials $g = 0$, and for world-fixed trials $g = 1$. Using this ratio, we are able to compute the expected eye excursion, $\hat{\varphi} = gd/m$, for any given translation distance even those we did not explicitly measure.

\subsubsection{Model}

Using a simple cue integration model, we investigated whether inter-subject and inter-condition differences in the observed PSEs in conditions containing a translation under free fixation depend on actual eye movement behavior. We modeled perceived distance, $p$, as a weighted linear combination of a vestibular and an oculomotor estimate of translation (\eqnref{p3:eq2}). We assumed that the vestibular estimate is equal to the actual translation, $m$, and that the oculomotor estimate is equal to expected eye movement given the  actual, $\hat{\varphi}_{m_i}$. As the weights represent the relative contributions of the oculomotor and vestibular systems, they can sum to any arbitrary value; in \eqnref{p3:eq2} they are fixed to 1. Thus, the weighting parameter $\alpha$ regulates the eye movement contribution and $1 - \alpha$  the vestibular contribution:

\begin{equation}
\label{p3:eq2}
p = \alpha \hat{\varphi}_m d + (1 - \alpha) m = \alpha g m + (1 - \alpha) m
\end{equation}

By definition, the probe displacement is perceived as equal in length to the 10 \si{\centi\metre} reference displacement at the PSE. By substituting both sides by the right hand side of \eqnref{p3:eq3} and using subscripts for reference ($r$) and probe intervals ($p$), we obtain:

\begin{equation}
\label{p3:eq3}
\alpha g_r m_r + (1 - \alpha) m_r = \alpha  g_p m_p \alpha + (1 - \alpha) m_p + \epsilon
\end{equation}

In the present experiment, the reference displacement, $m_r$, was always 10 \si{\centi\metre} and the probe displacement, $m_p$, was equal to the measured PSE for the presented combination of fixation types (i.e, $PSE$ in \eqnref{p3:eq1}). This model (i.e. \eqnref{p3:eq3}) was then fit to data from the body and world conditions using linear regression, finding weight α that minimizes the sum of squared errors ($\sum{\epsilon^2}$).

\begin{equation}
\label{p3:eq4}
m_r - m_p = \alpha(g_{f_p} m_p - g_{f_r} m_r + m_r - m_p) + \epsilon
\end{equation}

By only using data from conditions where a visual fixation point was present (i.e. body versus world) to fit the model, we examined if the same weight $\alpha$ also explains the PSEs found in the conditions where fixation was free. To this end, we solved Equation 3 for $m_p$ and computed PSE estimates, $P\hat{S}E$, for the body versus free and world versus free conditions (\eqnref{p3:eq5}).
                                                                                            \begin{equation}
\label{p3:eq5}
\hat{m}_p = P\hat{S}E = \frac
	{\alpha g_r + (1 - \alpha)}
	{\alpha g_p + (1 - \alpha)}
    m_r
\end{equation}

In addition to minimizing the sum of squared errors in \eqnref{p3:eq4}, we also fit \eqnref{p3:eq5} to the data in order to see if weight α depends on the way the model is formulated. Parameters obtained by fitting \eqnref{p3:eq5} fell well within the standard deviation reported in \tabref{p3:tab2} for all participants, suggesting that they did not depend on the way the model was formulated.

### 2015-03-21 src/paper3/p3_results.tex
Updated papers
Processed Luc and Pieters comments on Intro, P3 and P4

\begin{figure}
    \includegraphics[width=1.0\textwidth]{src/paper3/figure2.eps}

@@

The performance of one participant is illustrated in the top row of \figref{p3:fig2}. Each panel shows one main condition: body versus world fixation (red), world versus free fixation (green), and body versus free fixation (blue). The lighter and darker colors in each panel indicate which fixation type was the reference movement. The shift of the psychometric functions relative to the 10 \si{\centi\metre} reference (i.e. the PSE) quantifies the influence of fixation type. For example, the rightward shift of the light red curve in \figref{p3:fig2}A means that for a body fixation a longer translation ($\approx 19$ \si{\centi\metre}) was required for that translation to be perceived equivalent to a 10 \si{\centi\metre} reference translation with world fixation. On the other hand, the leftward shift of the dark red curve means that a shorter translation with world fixation ($\approx 7$ \si{\centi\metre}) was required for that translation to be perceived equivalent to the 10cm reference translation with body fixation. Together, these oppositely directed shifts demonstrate that translations with world fixation were perceived longer than equivalent translations with body fixation, regardless of which translation was the reference. Similarly, the shifts in \figref{p3:fig2}B shows that world fixation translations were also perceived to be longer than free fixation movements. \figref{p3:fig2}C shows that free fixation translations were perceived to be longer than body fixation translations.

Similar results were obtained for all subjects, as shown by the individual PSEs for all participants (bottom row of \figref{p3:fig2}. Statistical significance of the fixation-induced effects for each main condition (world versus body, \figref{p3:fig2}D; world versus free, \figref{p3:fig2}E; and free versus body, \figref{p3:fig2}F) were evaluated by comparing PSEs between the two reference conditions using a paired t-test. These PSEs were significantly different in all cases (world versus body, \ttest{7}{-4.09}{0.05}; world versus free, \ttest{7}{-2.48}{0.05}; free versus body, \ttest{7}{-3.38}{0.05}. As for the example subject, these results indicate that translations made with body fixation are perceived shorter than free fixation, which in turn are perceived shorter than the same translation with world fixation.

\begin{figure}
    \includegraphics[width=1.0\textwidth]{src/paper3/figure3.eps}

@@

In order to relate psychophysical performance to eye movement behavior we recorded and analyzed eye movements during both intervals of every trial for all subjects. Exemplar eye traces for the 10 \si{\centi\metre} reference translation for the three fixation types are depicted in \figref{p3:fig3}A. Although fixation behavior was quite accurate during both body and world fixation, catch-up saccades were typically observed during world fixation. Under free fixation, the amount of eye movement was intermediate between body and world fixation and behavior was more variable. A similar pattern was observed in all participants, as illustrated by the normalized eye movement data (see Methods, and \figref{p3:fig3}B).

\begin{figure}
    \includegraphics[width=0.5\textwidth]{src/paper3/figure4.eps}

@@

To evaluate the hypothesis that self-motion perception is modulated by eye movement behavior, we propose a linear model in which perceived translation is a weighted average of a vestibular estimate (equal to the actual translation) and an oculomotor estimate (equal to the normalized eye movement times the actual translation) (\eqnref{p3:eq2}). This model contains a single free parameter (), which corresponds to the weight given to the oculomotor estimate. We fitted this model to the two body versus world conditions and obtained the value of the oculomotor weight for every subject (\tabref{p3:tab2}). The average oculomotor weight is 0.25 \textpm0.12 (SD), indicating that the contribution of the eye movement signal to the self-motion estimate is about 25 percent. Note that participant 4, whose oculomotor weight is furthest from this mean (α = 0.06), also shows a radically different eye movement gain during the free-fixation (see \figref{p3:fig3}B). We then used these oculomotor weights along with the normalized eye movement values to predict the PSEs in the remaining four conditions according to \eqnref{p3:eq5}. The predicted PSEs are plotted against the actually observed PSEs in \figref{p3:fig4}. The positive correlation ($\rho = 0.78$, $p < 0.01$) between observed and predicted PSEs suggests that eye movements are indeed used in self-motion perception, even in the absence of a fixation point (i.e. during free fixation). Furthermore, the fact that data points generally cluster near the unity line shows that our simple model does reasonably well in predicting perceptual performance across subjects and conditions based on oculomotor weight and normalized eye movement magnitude only. This holds true even for subject 7 whose oculomotor weight (\tabref{p3:tab2}) was approximately double the average, yet whose data points remain close to the unity line.

>>>>

\begin{figure}
    \includegraphics[width=1.0\textwidth]{src/paper3/figure2.pdf}

@@

The performance of one participant is illustrated in the left column of \figref{p3:fig2}. Each row shows one main condition: body versus world fixation (top/red), world versus free fixation (middle/green), and body versus free fixation (bottom/blue). The lighter and darker colors in each panel indicate which fixation type was the reference movement (see Legend). The shift of the psychometric functions relative to the 10 \si{\centi\metre} reference (i.e. the PSE) quantifies the influence of fixation type. For example, the rightward shift of the light red curve in \figref{p3:fig2}A means that for a body fixation a longer translation (\about 19 \si{\centi\metre}) was required for that translation to be perceived equivalent to a 10\si{\centi\metre} reference translation with world fixation. On the other hand, the leftward shift of the dark red curve means that a shorter translation with world fixation (\about 7 \si{\centi\metre}) was required for that translation to be perceived equivalent to the 10 \si{\centi\metre} reference translation with body fixation. Together, these oppositely directed shifts demonstrate that translations with world fixation were perceived longer than equivalent translations with body fixation, regardless of which translation was the reference.  Similarly, the shifts in \figref{p3:fig2}B shows that world fixation translations were also perceived to be longer than free fixation movements and \figref{p3:fig2}C shows that free fixation translations were perceived to be longer than body fixation translations. Note that Figure 2 also shows an effects on slope, which will be further discussed in the paragraph \nameref{p3:sec:precision}.

Similar results were obtained for all subjects, as shown by the individual PSEs for all participants (right column of \figref{p3:fig2}). Statistical significance of the fixation-induced effects for each main condition (world versus body, \figref{p3:fig2}D; world versus free, \figref{p3:fig2}E; and free versus body, \figref{p3:fig2}F) was evaluated by comparing PSEs between the two reference conditions using a paired t-test. These PSEs were significantly different in all cases (world versus body, \ttest{7}{-4.09}{0.05}; world versus free, \ttest{7}{-2.48}{0.05}; free versus body, \ttest{7}{-3.38}{0.05}. As for the example subject, these results indicate that translations made with body fixation are perceived shorter than with world fixation, suggesting that self-motion perception is modulated by eye movements even in absence of full-field optic flow. The free fixation translations, which controlling for confounds of the small the fixation point, were perceived to be longer than body and shorter than world fixation translation intervals, which could be expected if their gain would be smaller than 1 but larger than 0.

\begin{figure}
    \includegraphics[width=1.0\textwidth]{src/paper3/figure3.pdf}

@@

\subsection{Eye movement contributions to self-motion perception}

In order to relate psychophysical performance to eye movement behavior we recorded and analyzed eye movements during both intervals of every trial for all subjects. Exemplar eye traces for the 10 \si{\centi\metre} reference translation for the three fixation types are depicted in \figref{p3:fig3}A. Fixation behavior was quite accurate for both body fixations, where no eye movements were expected, and world fixation, where eye movement excursions of \siang{11} were expected, seemingly supported by  catch-up saccades. Under free fixation, the amount of eye movement was intermediate between body and world fixation and behavior was more variable. A similar pattern was observed in all participants, as illustrated by the normalized eye movement data (see \nameref{p3:sec:methods}, and \figref{p3:fig3}B).

\begin{figure}
    \includegraphics[width=0.5\textwidth]{src/paper3/figure4.pdf}

@@

\begin{table}
    \begin{tabular}{llll}
    Participant & Parameter ($\alpha$) \textpm SD \\
    \hline
    1 & 0.27 (\textpm 0.04) \\
    2 & 0.27 (\textpm 0.05) \\
    3 & 0.35 (\textpm 0.04) \\
    4 & 0.06 (\textpm 0.04) \\
    5 & 0.13 (\textpm 0.03) \\
    6 & 0.33 (\textpm 0.04) \\
    7 & 0.58 (\textpm 0.02) \\
    8 & 0.21 (\textpm 0.02) \\    
    \end{tabular}

    \caption{Estimated eye movement contribution (α) to the perception of self-motion (see Equation 5). Standard deviations are based on a bootstrap for each participant.}

    \label{p3:tab2}
\end{table}

To quantify the role of eye movements in self-motion, we tested a linear model in which perceived translation is a weighted average of a vestibular estimate (equal to the actual translation) and an oculomotor estimate (equal to the normalized eye movement times the actual translation; \eqnref{p3:eq2}). This model contains a single free parameter ($\alpha$), which corresponds to the relative weight given to the oculomotor estimate. We fitted this model to the two body versus world conditions and obtained the value of the oculomotor weight for every subject (\tabref{p3:tab2}). The average oculomotor weight is 0.25 \textpm0.12 (SD), indicating that the contribution of the eye movement signal to the self-motion estimate is about 25 percent. Note that participant 4, whose oculomotor weight is furthest from this mean ($\alpha = 0.06$), also shows a radically different eye movement gain during the free-fixation (see \figref{p3:fig3}B). We then used these oculomotor weights along with the normalized eye movement values to predict the PSEs in the remaining four conditions according to \eqnref{p3:eq5}. The predicted PSEs are plotted against the actually observed PSEs in \figref{p3:fig4}. The positive correlation ($\rho = 0.78$, $p < 0.01$) between observed and predicted PSEs suggests that eye movements are indeed used in self-motion perception, even in the absence of a fixation point (i.e. during free fixation). Furthermore, the fact that data points generally cluster near the unity line shows that our simple model does reasonably well in predicting perceptual performance across subjects and conditions based on oculomotor weight and normalized eye movement magnitude only. This holds true even for subject 7 whose oculomotor weight (\tabref{p3:tab2}) was approximately double the average, yet whose data points remain close to the unity line.

\begin{figure}
    \includegraphics[width=0.5\textwidth]{src/paper3/figure5.pdf}

    \caption{Effect of difference movement amplitude between reference and probe interval (i.e. the bias) on response uncertainty ($\sigma$). A data point is shown for every participant and condition (symbol color) pair, following the same color scheme as in \figref{p3:fig2}. The dashed black line is the linear regression trend line.}

    \label{p3:fig5}
\end{figure}

\subsection{Precision depends on the PSE}
\label{p3:sec:precision}

The psychometric curves of the example participant in \figref{p3:fig2} show that precision ($\sigma^{-2}$ in \eqnref{p3:eq1}) decreases as the difference between translated distance in the reference and probe intervals (i.e. the bias) increases. To further investigate this effect, \figref{p3:fig5} shows a linear relation ($R^2 = 0.64$) between the bias and precision across all participants and conditions. This effect, which follows Weber’s perceptual law \cite{fechner1860} is consistent with the signal-dependence of (discrimination) precision that has been shown recently for vertical self-motion \cite{nesti2014}.

### 2015-03-21 src/paper3/paper3.tex
Updated paper 3
Processed Luc and Pieters comments on Intro, P3 and P4

\small {\bf Abstract} Self-motion is typically accompanied by compensatory eye movements that help minimize retinal slip and maximize dynamic visual acuity. To date, it is unknown whether these eye movements also have a reversed role, serving as a cue for self-motion perception. To address this question, we had participants ($n=8$) judge self-motion during different eye movement conditions in the absence of full-field optic flow.  In a 2-AFC task, participants indicated whether the second of two successive, passive, lateral whole-body translations was longer or shorter than the first. Eye movements during each translation were world-stationary, body-stationary or free in an otherwise dark room. Results show that the perceived translations were shorter with body-fixed gaze compared to world-fixed gaze. Using a linear model, we estimated the vestibular and eye movement contributions to self-motion perception and found that eye movement signals contribute approximately 25 percent. The model was validated independently by predicting the effects of eye movements on self-motion perception during the free eye movement conditions, taking the inter-subject variation in oculomotor weight and eye movement patterns into account. We conclude that eye movement signals influence self-motion perception, even in the absence of visual stimulation, and even when oculomotor and vestibular estimates are in conflict, e.g. during body-fixed gaze. We hypothesize that adverse consequences of this seemingly inflexible arrangement are minimal under natural conditions because eye movements and self-motion are highly correlated, and because eye movements are most often accompanied by veridical optic flow cues to self-motion.

@@

\noindent This chapter has been published as \newline
\noindent {\bf Clemens, I.A.H.}, Selen, L.P.J., MacNeilage, P.R. and Medendorp, W.P. \citeyear{clemens2015a}. Title. \emph{Journal}, volume(issue): from-to. \newline

>>>>

\small {\bf Abstract} Self-motion is typically accompanied by compensatory eye movements that help minimize retinal slip and maximize dynamic visual acuity. To date, it is unknown whether these eye movements also have a reversed role, serving as a cue for self-motion perception. To address this question, we had participants ($n=8$) judge self-motion during different eye movement conditions in the absence of full-field optic flow.  In a 2-AFC task, participants indicated whether the second of two successive passive lateral whole-body translations was longer or shorter than the first. Eye movements during each translation were world-stationary, body-stationary in an otherwise dark room. Results of these two conditions show that the perceived translations were shorter with body-fixed gaze compared to world-fixed gaze. Using a linear model, we estimated the relative contributions of vestibular and eye movement signals to self-motion perception and found that eye movement signals contribute approximately 25 percent. The model was independently validated by successfully predicting the effects of eye movements on self-motion in a third condition without any visual fixation, i.e. when the eyes were free to move. We conclude that eye movement signals influence self-motion perception, even in the absence of visual stimulation, and even when oculomotor and vestibular estimates are in conflict, e.g. during body-fixed gaze. We hypothesize that adverse consequences of this seemingly inflexible arrangement are minimal under natural conditions because eye movements and self-motion are highly correlated, and because eye movements are most often accompanied by veridical optic flow cues to self-motion.

@@

\noindent This chapter is being prepared for publication \newline
\noindent {\bf Clemens, I.A.H.}, Selen, L.P.J., MacNeilage, P.R. and Medendorp, W.P. \citeyear{clemens2015a}. %Title. \emph{Journal}, volume(issue): from-to. \newline

### 2015-03-21 src/paper4/p4_intro.tex
Processed Luc and Pieters comments on Intro, P3 and P4

When the eyes track world-centered objects, their angular displacement is directly related to the size of the motion of the observer \cite{schwarz1989, paige1988, mchenry2000, medendorp2002}. Because  the majority of fixations are on world-stationary  objects, we recently proposed that these tracking eye movements could also be used as a self-motion cue, in addition to optic flow and vestibular signals. To test this hypothesis, we compared self-motion perception in the absence of full-field optic flow during passively induced whole-body translations \cite{clemens2015a}. Supporting the notion that eye movement can act as a self-motion cue, our results showed that self-motion is underestimated during body-centered fixations (in which the eyes remain stationary in their orbits) compared to fixations on world-stationary objects (in which the eyes must move to maintain fixation).

Geometrically, eye movements that keep fixation on a world-centered target during lateral whole body translation (i.e. the linear vestibulo-ocular reflex; LVOR), must scale with fixation depth \cite{angelaki2004}. When fixating body-centered targets these eye movements must be suppressed irrespective of fixation distance \cite{angelaki2004}. Conversely, to serve as an adequate self-motion cue, the brain must internally scale the ensuing eye movement by fixation distance. Because we did not manipulate fixation distance in our previous study, it is unknown whether eye movements are used as a rudimentary cue for self-motion (i.e. without taking fixation depth i nto account), or are properly scaled in the mechanisms for self-motion perception.

>>>>

When the eyes track world-centered objects, their angular displacement is directly related to the size of the motion of the observer \cite{schwarz1989, paige1998, mchenry2000, medendorp2002}. Because  the majority of fixations are on world-stationary  objects, we recently proposed that these tracking eye movements could also be used as a self-motion cue, in addition to optic flow and vestibular signals. To test this hypothesis, we compared self-motion perception in the absence of full-field optic flow during passively induced whole-body translations \cite{clemens2015a}. Supporting the notion that eye movement can act as a self-motion cue, our results showed that self-motion is underestimated during body-centered fixations (in which the eyes remain stationary in their orbits) compared to fixations on world-stationary objects (in which the eyes must move to maintain fixation).

Geometrically, eye movements that keep fixation on a world-centered target during lateral whole body translation (i.e. the linear vestibulo-ocular reflex; LVOR), must scale with fixation depth \cite{angelaki2004}. When fixating body-centered targets these eye movements must be suppressed irrespective of fixation distance \cite{angelaki2004}. Conversely, to serve as an adequate self-motion cue, the brain must internally scale the ensuing eye movement by fixation distance. Because we did not manipulate fixation distance in our previous study, it is unknown whether eye movements are used as a rudimentary cue for self-motion (i.e. without taking fixation depth into account), or are properly scaled in the mechanisms for self-motion perception.

### 2015-03-21 src/paper4/paper4.tex
Updated paper 3, paper 4 and introduction'
Updated paper 4
Processed Luc and Pieters comments on Intro, P3 and P4

\chapter{}
\chaptermark{}

@@

\small {\bf Abstract}

@@

\noindent This chapter has been published as \newline
\noindent {\bf Clemens, I.A.H.}, Selen, L.P.J., MacNeilage, P.R. and Medendorp, W.P. \citeyear{clemens2014}. Title. \emph{Journal}, volume(issue): from-to. \newline

>>>>

% Incomplete compensation for fixation depth in eye movement driven self-motion perception

\chapter{Incomplete compensation for fixation depth for eye movement induced self-motion perception}
\chaptermark{}

@@

\small {\bf Abstract}
Previously, we have shown that the eye movements that typically accompany self-motion are used to augment self-motion perception. As fixation depth modulates the amplitude of these eye movements, it should be taken into account for a veridical oculomotor estimate of self-motion. To investigate whether the brain compensates for fixation depth, we had  participants ($n = 8$) judge self-motion during different eye movement conditions in the absence of full-field optic flow.  In a 2-AFC task, participants indicated whether the second of two successive passive lateral whole-body translations was longer or shorter than the first. At the start of each movement participants fixated either a nearby or a far away target, which was either body stationary or world stationary during the translation interval. Results show that the perceived translations were shorter for nearby world-fixed gaze compared to far away world-fixed gaze, indicating that eye movements are not properly scaled in self-motion perception. Together with the observation that self-motion perception is not affected by the depth of a body stationary fixation target, we conclude that eye movements are merely a rudimentary cue to self-motion, with a compensation for fixation depth that is partial at best.

@@

\noindent This chapter is being prepared for publication \newline
\noindent {\bf Clemens, I.A.H.}, Selen, L.P.J., MacNeilage, P.R. and Medendorp, W.P. \citeyear{clemens2015b}. %Title. \emph{Journal}, volume(issue): from-to. \newline
